---
title: "Area-to-area kriging with basis functions"
date: 2026-05-17
description: "Satellite observations are areal averages, not point measurements. Basis function methods reduce the intractable double-integral problem of area-to-area kriging to a single pass of cheap single integrals — provided the prior is chosen cleverly."
categories: [spatial statistics, downscaling, bayesian, kriging]
bibliography: reference.bib
---

In a [previous post](../kriging-gp-equivalence/index.qmd), I examine how kriging and Gaussian process regression can be formulated as Bayesian linear regression. The posterior mean is the kriging predictor.
This post is about what happens when observations are *areas* rather than points —
the area-to-area (ATA) kriging problem — and why the choice of basis functions is
not incidental but central to making the problem tractable.

## The naive approach is expensive

Classical point-referenced kriging predicts $v(s_0)$ from observations
$y_1, \ldots, y_n$ using the covariance structure of the field. When each observation
$y_i$ is actually an areal average over a footprint $D_i$,

$$y_i = \frac{1}{|D_i|}\int_{D_i} v(s)\,ds + \varepsilon_i$$

the covariance between two observations involves a *double integral* over each pair
of footprints:

$$\mathrm{Cov}(y_i, y_j) = \frac{1}{|D_i||D_j|}\iint_{D_i \times D_j} C(s,t)\,ds\,dt$$

With $n$ observations this requires $O(n^2)$ double integrals — one for every pair.
For $n = 10{,}000$ satellite soundings and footprints that may be kilometres across,
this is not just slow but often treated as a non-starter. That would require 100 million **double integrals** just to build the covariance matrix. Pretending each footprint is a point
at its centroid avoids the double integrals, but as we show below it miscalibrates
uncertainty in a systematic way.

None of the twelve methods in the @heaton2019case competition were designed for areal
observations. They are all point methods. This is not a critique — the competition
dataset happened to be on a regular grid with small effective footprints — but it
means the ATA problem is genuinely open for the irregular footprint settings common
in satellite remote sensing. However, the vast majority of the big-$n$ point kriging methods in the competition are also *basis function* methods, which, as I will argue, means they are likely extendable to areal data.

## Basis functions reduce $O(n^2)$ to $O(n)$

A basis function approach dissolves the double-integral problem. Represent the field
as $v(s) \approx \sum_k \beta_k \phi_k(s)$ and substitute:

$$y_i = \frac{1}{|D_i|}\int_{D_i} \sum_k \beta_k \phi_k(s)\,ds + \varepsilon_i
= \sum_k \beta_k \underbrace{\frac{1}{|D_i|}\int_{D_i} \phi_k(s)\,ds}_{A_{ik}} + \varepsilon_i$$

This gives the Bayesian linear model $y = A\beta + \varepsilon$ with prior
$\beta \sim \mathcal{N}(0, Q^{-1})$. Each entry $A_{ik}$ requires a *single* integral
of one basis function over one footprint — $O(nK)$ integrals total, where $K$ is the
number of basis functions. The cross-covariance structure between all pairs of
footprints is handled implicitly through the prior: the implied covariance between
$y_i$ and $y_j$ is $A_i Q^{-1} A_j^\top$, computed as matrix multiplication rather
than double integration.

This reduction from $O(n^2)$ double integrals to $O(nK)$ single integrals is not
free — it depends entirely on $Q$ being chosen to accurately represent the covariance
structure of the field. A poorly chosen prior produces a model that is computationally
cheap but statistically wrong. The choice of basis and prior is everything.

## What makes a good basis?

A desirable property is that the basis and prior $Q$ should correspond to a known and well-understood covariance model so that parameters have interpretable meaning and can be estimated from data. Secondarily, it is a very convenient property if the basis has closed-form integrals.

The SPDE approach of @lindgren2011explicit satisfies both. It begins from the
observation that a Matérn Gaussian random field is the stationary solution to the
stochastic partial differential equation

$$(\kappa^2 - \Delta)^{\alpha/2} v(s) = \mathcal{W}(s)$$

where $\Delta$ is the Laplacian, $\mathcal{W}(s)$ is Gaussian white noise, $\kappa$
controls the range, and $\alpha = \nu + d/2$ for smoothness $\nu$ in $d$ dimensions.
Discretizing this equation via finite elements on a triangular mesh produces piecewise
linear tent functions as the basis, and the precision matrix $Q$ is sparse and
derived directly from the SPDE operator — so it correctly represents a Matérn
covariance by construction. Lindgren also provides results for using indicator functions on a raster grid.

The single integrals $A_{ik} = \frac{1}{|D_i|}\int_{D_i} \phi_k(s)\,ds$ for tent
functions or indicator functions reduce to polygo intersection areas, computable exactly via
standard computational geometry. No quadrature, no approximation error, no dependence
on footprint resolution.

It is noteworthy that many other basis function types are easy to integrate. For instance, sinusoidal bases have closed-form integrals over triangulations, making periodic or Fourier bases feasible (but dense); while compactly supported polynomials such as used in LatticeKrig are also promisingly adapted to the areal setting.

## Why we'll start with it: SPDE won the satellite dataset

In the @heaton2019case competition--see [*my writeup*](../posts/kriging-competition/index.Rmd) for a summary--the SPDE/INLA method achieved the best RMSE on
the real satellite surface temperature data (1.53 RMSE) and was a top-performer on their simulation both in terms of speed and accuracy. Not only this, but their results hold for regular lattices, whose integrals are trivial to compute using area calculations available in common spatial packages like `sf`.

While it's a clear first-attempt to port to the areal setting, several methods are likely low-hanging-fruit as well. For example, another top performer for point-referenced data, LatticeKrig, uses a similarly motivated basis structure (with an intended spatially autoregressive prior) and would likely be easy to import to an areal setting. LatticeKrig uses basis functions are radial polynomials--which adds a step for integration beyond areal averages, but would not be challenging to integrate over triangulations.


## The connection to SAR models on regular grids

@lindgren2011explicit is primarily presented as a triangulation method, but they
also establish a clean result for regular grids that we will use. Using indicator functions as a basis is highly interpretable in terms of overlapping geometries over target regions.

On a regular grid with spacing $\delta$, the SPDE $(\kappa^2 - \Delta)^{\alpha/2}v = \mathcal{W}$ is discretized via finite differences rather than finite elements. For
$\alpha = 2$ (corresponding to $\nu = 1$ in two dimensions), the discrete Laplacian
at interior grid point $s_i$ is

$$\Delta v(s_i) \approx v(s_{i-1,j}) + v(s_{i+1,j}) + v(s_{i,j-1}) + v(s_{i,j+1}) - 4v(s_i)$$

Substituting into the SPDE gives

$$(\kappa^2 + 4)v(s_i) - \sum_{j \in N(i)} v(s_j) = \mathcal{W}(s_i)$$

which in matrix form is $Bv = \varepsilon$ where $B_{ii} = \kappa^2 + 4$ and
$B_{ij} = -1$ for rook neighbours. The precision matrix $Q = B^\top B / \sigma^2$
is exactly the precision of a **simultaneous autoregressive (SAR)** model with
rook neighbourhood structure. This is not an approximation — the finite-difference
discretization of the $\alpha=2$ SPDE *is* the rook SAR model.

The parameter correspondence is explicit. The SAR model has autoregressive
parameter $\rho$ and the SPDE has range parameter $\kappa$, related by

$$\kappa^2 = \frac{4(1-\rho)}{\rho\,\delta^2}$$

which gives the range in terms of $\rho$ and grid spacing $\delta$:

$$\text{range} = \sqrt{8/\kappa^2} = \delta\sqrt{\frac{2\rho}{1-\rho}}$$

This means fitting a SAR model and translating its $\rho$ back to a Matérn range
is exact for $\alpha=2$, and approximately valid for other smoothness values.

For the basis function representation on a regular grid the natural choice is
indicator functions — $\phi_j(s) = \mathbf{1}(s \in D_j)$, one per grid cell.
Then $A_{ik} = |D_i \cap D_k| / |D_i|$ is the area overlap fraction, computable
exactly via polygon intersection. The SAR precision $Q = B^\top B / \sigma^2$
is the correct prior for this basis under the SPDE connection.

## Simulation: does the area correction matter?

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE, comment="#>", echo=FALSE,
  fig.width=7, fig.height=4.5, dpi=150,
  message=FALSE, warning=FALSE
)
library(spatintegrate); library(fastblm); library(sf)
library(Matrix); library(spdep); library(ggplot2)
library(future); library(future.apply)

theme_post <- function()
  theme_minimal(base_size=12) +
  theme(panel.border=element_rect(fill=NA, colour="grey70"))

domain_bbox <- c(0, 0, 10e3, 10e3)
range_true  <- 3e3/3; sigma2_true <- 1.0; sigma_e <- 0.15
kappa_true  <- sqrt(8) / range_true

make_footprint <- function(cx, cy, w=1300, h=2300, deg=15) {
  a <- deg*pi/180; R <- matrix(c(cos(a),-sin(a),sin(a),cos(a)),2)
  v <- matrix(c(-w/2,-h/2,w/2,-h/2,w/2,h/2,-w/2,h/2,-w/2,-h/2),
              ncol=2,byrow=TRUE) %*% t(R)
  v[,1] <- v[,1]+cx; v[,2] <- v[,2]+cy; sf::st_polygon(list(v))
}
make_rook_W <- function(grid_sf) {
  nb<-spdep::poly2nb(grid_sf,queen=FALSE); n<-length(nb)
  i_o<-rep(seq_len(n),lengths(nb)); j_o<-unlist(nb,use.names=FALSE)
  n_nb<-lengths(nb); diag_idx<-which(n_nb<4L)
  Matrix::sparseMatrix(i=c(i_o,diag_idx),j=c(j_o,diag_idx),
    x=c(rep(1/4,length(i_o)),(4L-n_nb[diag_idx])/4),dims=c(n,n))
}
make_sar_Q <- function(grid_sf, rho)
  Matrix::crossprod(Matrix::Diagonal(nrow(make_rook_W(grid_sf)))-rho*make_rook_W(grid_sf))

fourier_fit_J <- function(J, fps_sf, domain_bbox, delta, rho_hat, phi_hat) {
  fg  <- fourier_freq_grid(J=J, domain_bbox=domain_bbox)
  A_f <- fourier_integrate_basis(fps_sf, fg)
  Q_f <- fourier_sar_Q_fun(freq_grid=fg, delta=delta)(
    c(logit_rho=qlogis(rho_hat), log_tau2=log(1)))$Q
  Qi  <- Matrix::Diagonal(x=1/Matrix::diag(Q_f))
  fit <- fit_fastblm(y, A_f, Q_f, phi=phi_hat, solver="woodbury", Q_inv=Qi)
  list(J=J, fg=fg, A=A_f, fit=fit, fitted=as.numeric(A_f %*% fit$posterior_mean))
}
```

```{r sim-field, include=FALSE, cache=TRUE}
set.seed(42)
grid_n <- 80; gx <- seq(0, 10e3, length.out=grid_n)
xy <- expand.grid(x=gx, y=gx)
D  <- as.matrix(dist(as.matrix(xy)))
Cov <- sigma2_true * (1 + kappa_true*D) * exp(-kappa_true*D)
L   <- chol(Cov + diag(1e-8, nrow(Cov)))
set.seed(7); xy$truth <- as.numeric(crossprod(L, rnorm(nrow(xy))))

n_obs <- 250
cx <- runif(n_obs, 2.5e3, 7.5e3); cy <- runif(n_obs, 2.5e3, 7.5e3)
fps_sf <- sf::st_sf(id=seq_len(n_obs),
  geometry=sf::st_sfc(mapply(make_footprint,cx,cy,SIMPLIFY=FALSE),crs=32619))
truth_sf <- sf::st_as_sf(xy, coords=c("x","y"), crs=32619)
mu_areal <- vapply(seq_len(n_obs), function(i) {
  ins <- sf::st_within(truth_sf, fps_sf[i,], sparse=FALSE)[,1]
  if (!any(ins)) NA_real_ else mean(xy$truth[ins])
}, numeric(1))
keep <- !is.na(mu_areal)
fps_sf <- fps_sf[keep,]; mu_areal <- mu_areal[keep]
y <- mu_areal + rnorm(sum(keep), sd=sigma_e); n_obs <- length(y)
```

```{r build-grid-A, include=FALSE, cache=TRUE}
domain_poly <- sf::st_as_sfc(
  sf::st_bbox(setNames(domain_bbox,c("xmin","ymin","xmax","ymax")),crs=32619))
sar_grid_sf <- sf::st_sf(
  cell_id=seq_along(sf::st_make_grid(domain_poly,n=c(80,80))),
  geometry=sf::st_make_grid(domain_poly,n=c(80,80)))
delta <- sqrt(as.numeric(sf::st_area(sar_grid_sf[1,])))
n_cells <- nrow(sar_grid_sf)

A_areal <- compute_overlap_fractions(soundings=fps_sf,fine_grid=sar_grid_sf,sparse=TRUE)

centroids <- sf::st_coordinates(sf::st_centroid(fps_sf))
cell_ids  <- sf::st_within(
  sf::st_as_sf(as.data.frame(centroids),coords=c("X","Y"),crs=32619),
  sar_grid_sf, sparse=FALSE)
centroid_cell <- apply(cell_ids, 1, function(r) if(any(r)) which(r)[1] else NA_integer_)
keep2 <- !is.na(centroid_cell)
A_point <- Matrix::sparseMatrix(
  i=seq_len(n_obs)[keep2], j=centroid_cell[keep2],
  x=rep(1,sum(keep2)), dims=c(n_obs,n_cells))
```

```{r fit-models, include=FALSE, cache=TRUE}
future::plan(future::multisession, workers=5L)
Q_fun_sar_ml <- function(theta) {
  rho <- plogis(theta[["logit_rho"]]); Q <- make_sar_Q(sar_grid_sf, rho)
  list(Q=Q, log_det_Q=as.numeric(Matrix::determinant(Q,logarithm=TRUE)$modulus))
}
fit_ml <- tune_ml(y=y, A=A_areal, Q_fun=Q_fun_sar_ml,
  theta_init=c(logit_rho=qlogis(0.95)),
  lower=c(logit_rho=qlogis(0.5)), upper=c(logit_rho=qlogis(0.999)), verbose=FALSE)
rho_hat <- plogis(fit_ml$theta[["logit_rho"]]); phi_hat <- fit_ml$phi
kappa2_hat <- 4*(1-rho_hat)/(rho_hat*delta^2); range_hat <- sqrt(8/kappa2_hat)

Q_fit <- make_sar_Q(sar_grid_sf, rho_hat)
Q_inv_fit <- function(v) as.numeric(Matrix::solve(Q_fit, v))
fit_areal <- fit_fastblm(y,A_areal,Q_fit,phi=phi_hat,solver="woodbury",Q_inv=Q_inv_fit)
fit_point <- fit_fastblm(y,A_point,Q_fit,phi=phi_hat,solver="woodbury",Q_inv=Q_inv_fit)
fitted_areal <- as.numeric(A_areal %*% fit_areal$posterior_mean)
fitted_point <- as.numeric(A_point %*% fit_point$posterior_mean)

se_areal <- posterior_se(fit_areal, A_new=as.matrix(A_areal))
se_point <- posterior_se(fit_point, A_new=as.matrix(A_point))
z90 <- qnorm(0.95)
covered_areal <- abs(mu_areal - fitted_areal) < z90 * se_areal
covered_point <- abs(mu_areal[keep2] - fitted_point[keep2]) < z90 * se_point[keep2]
```

We simulate a Matérn($\nu=1$) field with range 3 km over a 10×10 km domain, then
generate 250 satellite-style footprint observations — roughly 1.3 km × 2.3 km
parallelograms at 15° rotation, representative of a low-Earth-orbit sounder. We fit
two versions of the same SAR-prior model:

- **Areal model:** $A_{ik}$ = area overlap fraction of footprint $i$ and cell $k$ —
  the correct observation model.
- **Point model:** $A_{ik} = 1$ for the single cell containing the footprint centroid,
  0 elsewhere — the standard approximation used by all Heaton methods.

Both use the same prior (tuned via marginal likelihood on the areal model) and the
same noise variance $\phi$.

```{r plot-setup, fig.height=4}
ggplot() +
  geom_raster(data=xy, aes(x,y,fill=truth)) +
  geom_sf(data=fps_sf, fill=NA, colour="white", linewidth=0.3, alpha=0.5) +
  scale_fill_distiller(palette="RdYlBu") +
  labs(title=sprintf("Matérn field (range = %.0f km) with %d footprints",
                     range_true/1e3, n_obs),
       x="Easting (m)", y="Northing (m)", fill="v(s)") +
  coord_sf() + theme_post()
```

```{r plot-coverage-rate}
cov_df <- data.frame(
  model    = c("Areal (correct)", "Point centroid"),
  coverage = c(mean(covered_areal), mean(covered_point)),
  nominal  = 0.90
)
ggplot(cov_df, aes(model, coverage, fill=model)) +
  geom_col(width=0.5, show.legend=FALSE) +
  geom_hline(yintercept=0.90, linetype=2) +
  scale_y_continuous(limits=c(0,1), labels=scales::percent) +
  scale_fill_manual(values=c("Areal (correct)"="steelblue","Point centroid"="firebrick")) +
  labs(title="90% posterior credible interval coverage",
       subtitle="Dashed line = nominal 90%",
       x=NULL, y="Coverage") +
  theme_post()
```

The point approximation systematically miscalibrates uncertainty. Footprints
spanning regions of high spatial variability are most affected — the observation
model treats a 3 km footprint as if it were a point, ignoring the spatial averaging
that reduces the effective noise. The areal model, because it correctly propagates
the footprint geometry into $A$, gives calibrated intervals at the nominal rate.

## Swapping to a Fourier basis: free given the SAR fit

Once the SAR parameters $(\rho, \phi)$ are estimated, there is a natural alternative
basis that inherits the same covariance structure: the Neumann cosine functions

$$\phi_{\mathbf{j}}(s) = \prod_{k=1}^2 \sqrt{2-\delta_{j_k,1}}\,
\cos\!\left(\frac{(j_k-1)\pi s_k}{L_k}\right)$$

These diagonalize the SAR precision matrix exactly — the prior $Q$ is diagonal in this
basis, with eigenvalues given analytically by the SAR parameter $\rho$:

$$Q_{\text{Fourier},\mathbf{j}} = \frac{(1-\rho\lambda_{W,\mathbf{j}})^2}{\tau^2}$$

where $\lambda_{W,\mathbf{j}}$ are the eigenvalues of the rook weight matrix $W$.
No re-estimation is needed: $\rho$ and $\phi$ translate directly from the SAR fit.

The key advantage for ATA kriging is that the integrals $A_{ik} = \frac{1}{|D_i|}
\int_{D_i} \cos(\omega_x s_x)\cos(\omega_y s_y)\,ds$ have exact closed forms via the
identity $\cos(A)\cos(B) = \frac{1}{2}[\cos(A-B)+\cos(A+B)]$ and the complex
exponential triangle formula. Like the SPDE/INLA tent function integrals, these are
exact — no quadrature error, no dependence on footprint resolution.

The Fourier basis also decouples prediction resolution from fitting resolution. The
SAR raster model is tied to an 80×80 grid; the Fourier model with the same fitted
parameters predicts at any resolution by evaluating the basis functions at new
locations, with no refitting required.

```{r fit-fourier, include=FALSE, cache=TRUE}
# Parameter matching: get kappa2 and phi for Fourier basis
# h_norm = normalised cell size on [0,1]^2
h_norm      <- 1 / round(sqrt(n_cells))   # = 1/80 for 80x80 grid
L           <- max(diff(domain_bbox[c(1,3)]), diff(domain_bbox[c(2,4)]))
J           <- round(L / delta)
 
params      <- get_matern_parameters_from_SAR(rho_hat, sar_precision=1, h=h_norm)
KAPPA2_NORM <- params$kappa2
phi_f       <- phi_hat * params$matern_variance
 
# Frequency grid: unit-square convention, top K modes sorted by eigenvalue
K       <- min(10000L, J^2L)
bbox_sq <- domain_bbox   # already square
 
kx_all     <- rep(0:J, each=J+1L)
ky_all     <- rep(0:J, times=J+1L)
lambda_all <- pi^2 * (kx_all^2 + ky_all^2)
ord        <- order(KAPPA2_NORM + lambda_all)
kx         <- kx_all[ord[seq_len(K)]]
ky         <- ky_all[ord[seq_len(K)]]
 
freq_grid <- list(
  omega_mat   = cbind(omega_x = kx * pi / L, omega_y = ky * pi / L),
  norm_const  = ifelse(kx==0L, 1, sqrt(2)) * ifelse(ky==0L, 1, sqrt(2)),
  indices     = cbind(j1 = kx+1L, j2 = ky+1L),
  J1 = max(kx)+1L, J2 = max(ky)+1L,
  domain_bbox = bbox_sq, norm = "unit_square"
)
 
# Design matrix and prior
A_f         <- fourier_integrate_basis(fps_sf, freq_grid)
lambda_norm <- pi^2 * (kx^2 + ky^2)
q_diag      <- (KAPPA2_NORM + lambda_norm)^2
Q_f         <- Matrix::Diagonal(x = q_diag)
 
# Fit
fit_fourier <- fit_fastblm(
  y      = y,
  A      = A_f,
  Q      = Q_f,
  phi    = phi_f,
  solver = "woodbury",
  Q_inv  = function(v) v / q_diag
)
 
fitted_fourier <- as.numeric(A_f %*% fit_fourier$posterior_mean)

```

```{r plot-multiresolution, fig.height=8, cache = TRUE}
# Build sf grids at three resolutions for prediction
make_grid_sf <- function(n, bbox, crs=32619) {
  sf::st_sf(geometry = sf::st_make_grid(
    sf::st_as_sfc(sf::st_bbox(
      setNames(bbox, c("xmin","ymin","xmax","ymax")), crs=crs)),
    n = c(n, n), square = TRUE
  ))
}
 
grids_sf <- list(
  "Coarse (40x40)"  = make_grid_sf(40,  domain_bbox),
  "Medium (80x80)"  = make_grid_sf(80,  domain_bbox),
  "Fine (200x200)"  = make_grid_sf(200, domain_bbox)
)
 
pred_df <- do.call(rbind, Map(function(nm, g) {
  A_pred <- fourier_integrate_basis(g, freq_grid)
  pred   <- as.numeric(A_pred %*% fit_fourier$posterior_mean)
  cents  <- sf::st_coordinates(sf::st_centroid(g))
  data.frame(x = cents[,1], y = cents[,2], pred = pred, res = nm)
}, names(grids_sf), grids_sf))
 
ggplot(pred_df, aes(x, y, fill = pred)) +
  geom_raster() +
  facet_wrap(~res, ncol = 1) +
  scale_fill_distiller(palette = "RdYlBu") +
  coord_equal() +
  labs(title = "Fourier-SAR posterior mean at three resolutions",
       subtitle = "Same fitted coefficients — no refitting",
       x = "Easting (m)", y = "Northing (m)", fill = "v̂(s)") +
  theme_post()
```

The three panels show predictions at 40×40, 80×80, and 200×200 resolution from
the same fitted model. Adding resolution costs only basis function evaluations,
not model re-estimation (though there may be some open questions as to what resolution of SAR model is necessary to get good estimates of these parameters, they still inherit properties of whatever procedure is used *for the SAR model*, such guarantees from maximum likelihood).

## Summary

The ATA problem reduces from $O(n^2)$ double integrals to $O(nK)$ single integrals
whenever the covariance structure is encoded in a prior $Q$ on basis coefficients
rather than built directly from $C(s,t)$. Two things are convenient for this to work:
the basis functions must be integrable over footprints in closed form (or via quadrature), and $Q$
must adequately represent the target covariance.

The SPDE framework of @lindgren2011explicit satisfies both conditions for Matérn
fields — by construction, since $Q$ is derived directly from the SPDE operator. On
a regular grid this recovers the SAR model exactly, establishing a clean parameter
correspondence with classical lattice statistics. The Fourier basis inherits this
correspondence and adds resolution flexibility, at the cost of a dense rather than
sparse $A$ matrix.

| | Areal SAR | Point SAR | Fourier-SAR |
|---|---|---|---|
| **Observation model** | Correct | Approximate | Correct |
| **Integration** | Exact (overlap fractions) | None (centroid) | Exact (analytic) |
| **Coverage** | Calibrated | Miscalibrated | Calibrated |
| **Resolution** | Fixed grid | Fixed grid | Arbitrary |
| **Prediction at new resolution** | Refit | Refit | Free |
| **Prior** | SAR / Matérn | SAR / Matérn | Diagonal / Matérn |
